% !TEX program = pdflatex
% Supplementary Material for Mamba3Yolo
\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{times}
\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{caption}
\usepackage{listings}
\usepackage{microtype}

\title{Supplementary Material\\[0.3em]
\large Mamba3Yolo: Mamba-3 Enhanced Real-Time Object Detection\\
with Explainability and Edge Quantization}
\author{Anonymous / Authors}
\date{}

\begin{document}
\maketitle

\section{Additional Implementation Details}

\subsection{Mamba3ODSSBlock Hyper-parameters}
Default settings used in all experiments (Tiny scale):

\begin{center}
\begin{tabular}{lc}
\toprule
Hyper-parameter & Value \\
\midrule
$d_{\text{state}}$ & 64 \\
expand ratio & 2 \\
headdim & 64 \\
MIMO rank $r$ & 4 \\
DropPath rate & 0.0 -- 0.1 (linearly increasing with depth) \\
LSBlock kernel & $3\times3$ depthwise \\
RGBlock expansion & 2.0 \\
\bottomrule
\end{tabular}
\end{center}

\subsection{Training Recipe}
\begin{itemize}[leftmargin=*]
  \item Optimizer: AdamW ($\beta_1=0.9$, $\beta_2=0.999$)
  \item Initial LR: $1\times10^{-3}$, cosine decay to $1\times10^{-5}$
  \item Weight decay: 0.05
  \item Epochs: 300 (COCO), 100--150 (medical multi-domain)
  \item Batch size: 16--64 (depending on GPU memory)
  \item Augmentations: Mosaic (prob 1.0), MixUp (0.1), random affine, HSV, horizontal flip
  \item EMA decay: 0.9999
  \item Loss: CIoU + DFL + BCE (classification) with Task-Aligned Assigner (final version)
\end{itemize}

\subsection{Quantization Protocol}
\begin{enumerate}[leftmargin=*]
  \item Collect activation statistics on 256 images from the validation set.
  \item Apply per-token static scaling for Mamba-3 layers (following PTQ4VM).
  \item SmoothQuant-style migration of outliers for Linear layers.
  \item Export to ONNX (opset 17) then build TensorRT engine with \texttt{--fp16} or \texttt{--int8} + calibrator.
\end{enumerate}

\section{Additional Ablations}

\subsection{Effect of Scan Directions}
We tested 1, 2, and 4 directional scans inside Mamba3SS2D. Four directions give the best AP$_s$ (+0.7 over two directions) at a modest latency cost (+4\,\%).

\subsection{MIMO Rank Sensitivity}
\begin{center}
\begin{tabular}{lccc}
\toprule
MIMO rank $r$ & mAP & Params & FPS \\
\midrule
1 (SISO) & 45.3 & 6.1M & 191 \\
2        & 45.6 & 6.4M & 188 \\
4 (default) & \textbf{45.9} & 6.8M & 185 \\
8        & 46.0 & 7.5M & 179 \\
\bottomrule
\end{tabular}
\end{center}

\section{Extra Qualitative Results}

\subsection{Failure Cases}
Typical failure modes remaining after Mamba-3 upgrade:
\begin{itemize}[leftmargin=*]
  \item Extremely small objects ($<8\times8$ px) under heavy occlusion
  \item Motion blur in endoscopic sequences
  \item Domain shift to completely unseen imaging devices (different color response)
\end{itemize}

\subsection{Additional Grad-CAM++ Examples}
(Insert high-resolution figures generated by \texttt{scripts/generate\_xai\_figures.py} here.)

\begin{center}
\textit{[Figure S1: Grad-CAM++ on polyp, blood-cell and fundus images]}
\end{center}

\section{Broader Impact \& Limitations}
Mamba3Yolo is intended for research and clinical decision support, not as a standalone diagnostic device. All medical results should be validated under local regulatory frameworks before deployment. The model may inherit biases present in the public training sets (e.g., under-representation of certain skin tones or equipment vendors). We encourage users to fine-tune on local data and to keep a human expert in the loop.

\section{Reproducibility Checklist}
\begin{itemize}[leftmargin=*]
  \item Code: \url{https://github.com/ShMazumder/Mamba3Yolo}
  \item Configs: \texttt{configs/models/}, \texttt{configs/datasets/}
  \item Exact package versions: \texttt{requirements.txt}
  \item Random seed: 42 (all experiments)
  \item Hardware for reported numbers: NVIDIA T4 (Kaggle), RTX 3090, Jetson Orin NX
\end{itemize}

\section{Societal Impact Statement}
Improved real-time, explainable detectors can lower the barrier for AI-assisted screening in low-resource clinics. At the same time, over-reliance on automated systems without proper validation can lead to missed diagnoses. We release the full pipeline (including quantization and XAI) so that the community can audit, improve, and adapt the method responsibly.

\end{document}
